Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-14 paper/manuscript.tex
a few small edits

But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as ``shifts'' in a site's amino-acid preferences.
Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.

@@

\subsection{Two Env homologs from clade A transmitted-founder viruses}
The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic,ronen2015hiv}.

@@

\subsection{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
This last consideration is especially important when comparing mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
Our modified the deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

@@

While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B, compare red and blue versus gray plots).
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the two Envs.
Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.

@@

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters, with the re-scaling done as described in \citet{hilton2017phydms}.
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.

@@

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
\caption{\label{fig:divsel}
Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
{\bf (A)}
One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites are evolving slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
{\bf (B)}
Sites of slower or faster than expected evolution given the BF520 experiments.
For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
{\bf (C)}
The results are similar regardless of whether the BG505 or BF520 experiments are used.
Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
{\bf (D)} 
A large cluster of sites that evolve slower than expected is likely involved in Env's transition between open and closed conformational states.
Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form a hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.
}
\figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they fall into neither of these categories.
Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
\figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
The $Q$-value indicated is the maximum of that for the two homologs.
}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
\end{figure}

>>>>

But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.
Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.

@@

\subsection{Two Envs from clade A transmitted-founder viruses}
The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic,ronen2015hiv}.

@@

\subsection{Deep mutational scanning of each Env}
We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
Our modified the deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

@@

While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B, compare red and blue versus gray plots).
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.

@@

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.

@@

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
\caption{\label{fig:divsel}
Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
{\bf (A)}
One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to gray to red based on the strength of evidence that sites are evolving slower than expected (blue), as expected (gray) or faster than expected (red) given the BG505 experiments.
Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
{\bf (B)}
Sites of slower or faster than expected evolution given the BF520 experiments.
For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
{\bf (C)}
The results are similar regardless of whether the BG505 or BF520 experiments are used.
Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
{\bf (D)} 
A large cluster of sites that evolve slower than expected is likely involved in Env's transition between open and closed conformational states.
Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form a hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.
}
\figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} Envs, or whether they fall into neither of these categories.
Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
\figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both Envs.
Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
For each site, the plots show the preferences averaged across replicates and re-scaled for each Env, as well as the frequencies of amino acids in the clade A Env alignment. 
The $Q$-value indicated is the maximum of that for the two Envs.
}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
\end{figure}
